---
title: "group project"
author: "Cody, Anna,Aadarsh,Nick"
date: "2025-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Project Proposal
The project proposal should be a knitted R Markdown document which contains the following elements.

The names of the students in the group.
Cody, Anna, Aadrsh, Sankar, Nick 

A brief description of how the group will communicate and work together.

we use github to make a github shared space, and we use terminal, git pull and git push to recieve and send the latest of our work. Anna will be working on the which topic The question or questions of interest that motivate the planned study. part and Cody will be working on finding the sources of the data. Aadrsh will be making the plot and Nick is working on analyzing the variables. Also we will contact each other on Instagram trough a group chat.

When during the week can the group meet?
On Monday 2:30 after class
Consider needs of all group members.

What platforms will the group use to communicate and share documents.

we will use github to share documents and use Instagram group chat to communicate. 

The question or questions of interest that motivate the planned study.

1.How have patterns of crime in Los Angeles changed from 2020 to 2025 across different geographic areas, times, and types of offenses?
2. How has the transition to the new NIBRS-based reporting system in 2024 affected the accuracy, frequency, or completeness of reported crime data in Los Angeles?

Who collected the data?

Cody


How were the data collected?
```{r}
x<-read.csv("../stat240/Crime_Data_from_2020_to_Present.csv")
x





```
the data is collected through the public safety NIBRS website and it is downloaded into my desktop stat240 file I both put this file and the group project Rmd file in this space. so i can use read.csv to import my dataset.



What are key variables in the data set, what do they measure, and how are they related to your key questions?
Key variables include the dates and times the crimes occurred, the area (including long and lat), as well as the crime types. 

What does a single row of the data set represent in real terms?
A single row represents an official crime reported by the LAPD 

If the data may be thought of as a sample from some larger population, what is that larger population?
The larger population could include unreported crimes in LA, or even all crimes across the world in general.

Note: if you get the data from a data repository such as Kaggle, you will need to dig into supplementary information to answer these questions.
It is insufficient to simply report a link to a Kaggle repository.

Include a description of a graph which will show how the data informs the primary question of interest.
One possible graph could be a time series plot showing monthly crime counts for major offense categories from 2020–2025.

Describe what variables you will include and what type of graph you will create.
DATE.OCC and TIME.OCC, as well as Crm.Cd.Desc.

Include a description of a possible model for the data and associated methods of inference or prediction.
One possible model would predict the number of crimes per month as a function of the date. This would let us predict the number of crimes in the future.
We could also create a model to see if there’s correlation between the weapons used and the victim’s features (age, sex, descent).

```{r monthly_trend, echo=TRUE}
library(tidyverse)

x <- read.csv("Crime_Data_from_2020_to_Present.csv")
x$DATE.OCC <- as.Date(x$DATE.OCC, format="%m/%d/%Y")

monthly <- x %>%
  mutate(month = as.Date(format(DATE.OCC, "%Y-%m-01"))) %>%
  count(month)

ggplot(monthly, aes(x = month, y = n)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Monthly Crime Trends in Los Angeles (2020–2025)",
    x = "Month",
    y = "Total Reported Crimes"
  ) +
  theme_minimal()
```

This line graph indicates the amount of overall crimes reported in Los Angeles, over time, between 2020 and 2025, with every point representing the total amount of crimes reported in one month, The trend line indicates whether overall crime has increased, decreased or remains stable within those years. This visual assists in answering our main research question relating to how the crimes have fluctuated over time. It also sets the stage for our secondary question regarding the impact of the 2024 NIBRS reporting transition - Should there be an increased or decreased spike after early 2024, this may indicate a relation to the nuance of the new reporting system.